Adaptive Waveform Inversion for Transmitted Wave Data

### 2024-11-11 repro/papers/awi/paper.tex
more cleanup

\begin{abstract}
Adaptive Waveform Inversion
applied to transient transmitted wave data yields estimates of index of
refraction (or wave velocity) similar to those obtained by travel time
inversion, provided that the data contain a single smooth wavefront.
\end{abstract}

\section{Introduction}
Full waveform inversion (FWI) refers to the inference of mechanical
parameter distribution within a material object  (the earth, a human
body, a manufactured artifact,...) from remote measurements of waves
propagating through it. These mechanical parameters appear as
coefficients in systems of partial differential equations governing
wave propagation, and numerical solutions of these wave equations can
be used to predict experimental data {\em ab initio}. FWI consists in
updating the parameters to minimize a numerical measure of  the difference between computationally
predicted wave data and actual observed
data. This data is typicallyl oscillatory. If wave velocities or their parametric equivalents are amongst the parameters
to be estimated, then updates cause shifts in the arrival times of
data oscillations, which in turn causes commonly used numerical error measures
to oscillate. This ``cycle skipping'' phenomenon may cause iterative local optimization algorithms (relatives of
Newton;'s method, the only practical algorithmic choices) to stagnate
far from good data fit, unless the initial
parameter estimates accurately predict the arrival times of waves \cite[]{GauTarVir:86,VirieuxOperto:09,HuetalIEEE:18}.

@@

This paper supplies such an explanation for a special case of AWI, the
recovery of smooth acoustic medium parameters from data generated at
localized energy sources, propagated through a fluid, and recorded at
receiver locations remote from the sources (transmission data). I show that the AWI
objective function is an asymptotic (in time frequency) approximation
to the mean-square time-of-travel misfit, provided that arrival times
of transmitted waves are single. That is, a unique ray of geometric
asymptotics connects each sources-receiver pair, as is the case if the associated
Riemannian metric is simple in a domain containing the sources and
receivers \cite[]{PestovUhlmann:05,StefanovUhlmann:05}. The error in
this approximation is proportional to the source mean-square
wavelength. I also show that this relation fails if arrival times have
higher multiplicity. Travel time tomography (minimization of
mean-square travel time error) appears to be much less sensitive to
choice of initial model than is FWI. In fact, it is commonly used to
construct an initial model for FWI
\cite[]{Bordingetal:87,SirguePratt:04,VirieuxOperto:09}. At least for
acoustic transmission data with single arrivals, AWI implicitly
combines travel time tomography with FWI.

@@

  \begin{itemize}
  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
    L^2(\bR)$;
  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
    \lambda_*\le 1$ depending on
    $M, X_{sr}$ so that
    \begin{itemize}
    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$.
    \end{itemize}
  \end{itemize}
  Choose $r > 0$ and assume that
  \begin{equation}
    \label{eqn:lamsig}
    \sigma = r \lambda.
  \end{equation}.
  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
  \rightarrow \bR^+$ with $C_* \le
  W \le C^*$ so that for
  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{itemize}
  \item[2. ] for $0 <  \lambda \le 1$, 
    \begin{equation}
      \label{eqn:mswitt}
      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} (W[\kappa^*,\kappa](\bx_s,\bx_r))^{\frac{1}{2}}|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
      \lambda^{\frac{1}{2}};
    \end{equation}
  \item[3. ]for sufficiently small $\lambda$, 
    \begin{equation}
      \label{eqn:awitt}
      \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
        |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
    \end{equation}
  \end{itemize}
\end{theorem}
\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
2b show that for this to be the case, the regularization parameter $\sigma$
must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
would clearly suffice, however the argument lis slightly cleaner if
$\sigma$ is simply held proportional to $\lambda$ as $\lambda
\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.

Summation over the source-receiver positions in $X_{sr}$ yields
\begin{cor}
  \label{thm:cor1}
  In addition to the assumptions of Theorem \ref{thm:main}, define $J_{\rm
    AWI,\sigma}$ and $J_{\rm MSWI,\sigma}$ by equations
  \ref{eqn:awidefmain}, \ref{eqn:mswidefmain} respectively. Then
\begin{equation}
\label{eqn:cor1mswi}   
\lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa]^2(\bx_s,\bx_r) |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
\end{equation}
and
\begin{equation}
\label{eqn:cor1awi}   
\lim_{\lambda \rightarrow 0} J_{\rm AWI,r\lambda}[\kappa;\dl]
= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
\end{equation}
\end{cor}

@@

In fact, the behaviour inherited from travel-time inversion in the
single arrival case, namely the absence of spurious stationary points
(``cycle-skipping''), generally fails when multiple arrivals with
significant energy are present in the data. \cite{Symes:94c}
illustrated this phenomenon for a version of MSWI. Crosswell
source-receiver geometry often generates multiple arrivals,
particularly near low-velocity sedimentary layers that are natural
targets of seismic exploration and act as wave
guides. \cite{Plessix:00} successfully applied a version of MSWI to
crosswell waveform inversion by removing traces with close source and
receiver depths, thus leaving only data that tended to be dominated by
single arrivals. \cite{HuangSymes:Geo17} and
\cite{SymesChenMinkoff:24} illustrate the failure of the
version of MSWI reported here to avoid cycle-skipping in inversion of
strongly refracting models, using frequency-domain implementation of
the basic definition \ref{eqn:mswidefmain} and time-domain implementation
of the reduced penalty function \ref{eqn:mswipenred} respectively. \cite{Yongetal:GJI23}
have shown that precisely the same issue afflicts AWI, and have
proposed a remedy by localizing the definition of the objective to
short time intervals, within which multiple wavefront arrrivals are
unlikely. \cite{Symes:23} constructs an extension based on surface
sources, as opposed to the source-receiver extension which is the
basis of MSWI and AWI. This surface source extension generates an
invertible map from extended source to data.

@@

Of courese, the adaptive filter just computed is not square
integrable, and must be regularized for use in a square-integral
construction like that of AWI or MSWI. That amounts essentially to
replacing the $\delta$s in the above formulae by a square-integrable
approximate Dirac function, of roughly a wavelength in width. In the
preceding section, this approximate Dirac was $g$, so use the same
notation here. The regularized version of $u(t)$ (solution of the
least squares problem \ref{eqn:reg}) contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
and $g(t-(\tau_1-\tau^*_0))$. These are approximately scaled by
$\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
definitions \ref{eqn:mswidefmain} and \ref{eqn:awidefmain} of the MSWI and AWI
objectives. For other choices of objective, they contribute similarly
to the total. They will eventually begin to cancel as $m$ approaches
$\kappa^*$, however this cancellation happens only when the times are
within the width of the approximate 
Dirac waveform $g$ - that is, within a wavelength.

This argument shows that when multiple arrivals carry significant energy, any
inversion approach based on the source-receiver extension, similar to
AWI or MSWI, will cycle-skip, that is, change rapidly when the
estimated model approaches its optimal value.

It should be emphasized that this phenomenon is primarily a feature of the
source-receiver extension, rather than of the specific objective
function defined in terms of it. In that respect, the source-receiver
extension is similar to the common surface parameter extensions that
have been studied for inversion of reflection data. Inversion methods
based on these extensions also fail to avoid cycle-skipping in the presence of
geometrical ray complexity \cite[]{geoprosp:2008}.

@@

The reader will
note that a normalization component of the AWI objective function
definition is important in reaching this conclusion. The connection
between objective normalization and the asymptotic equivalence to
travel time inversion for transmission data has nothing to do with the
reasons for normalizing the objective previoiusly advanced in the
literature \cite[]{Warner:16,LiAlkhalifah:21}.

@@

Some of these constraints
could certainly be relaxed. For instance, exponential decay of local
energy implies that the finite time duration of actual measured data
can be treated as a perturbation of the infinite time data setting
posed here. The conclusions of this paper likely hold in some form
even for non-smooth sets of models, provided that the data for such models is
dominated by ballistic transmitted waves. The exception is the single-arrival hypothesis: it is
essential, as without it, the link between the travel time and
extended waveform inversion is lost. This is a feature of the
source-receiver extension, not of a particular objective function
(MSWI, AWI,...) formulated in terms of it. This extension treats each receiver signal
independently. If only a single identifiable waveform arrives at
each receiver, AWI (and similar approaches) can implicitly identify
its arrival time, and the link to travel time inversion is
established. If distinct waveforms are present at multiple times,
the single receiver signal does not determine which waveform in the
predicted signal should align with which waveform in the data to be
fitted. This ``crosstalk'' ambiguity is analyzed in the fifth
section, and afflicts all approaches based on the source-receivre
extension.

The analysis presented here has nothing to say about AWI applied to
reflected wave data. \cite{Warner:16} and others have
suggested that AWI is effective in that context. If so, the underlying
mechanism remains to be explained.

>>>>

\begin{abstract} Adaptive Waveform Inversion (AWI) applied to
  transient transmitted wave data can yield estimates of index of
  refraction (or wave velocity) similar to those obtained by travel
  time inversion. The AWI objective function measures normalized
  mean-square dispersion about zero time of a family of filters, one
  filter for each source-reeciver pair, designed to match predicted
  data to observed data. Provided that the data contain a single
  smooth wavefront, this function approaches the mean square
  traveltime error as data wavelength tends to zero. The scaling of
  each filter to have unit norm is responsible for the simple relation
  with travel time tomography. A similar approach, Matched Source
  Waveform Inversion (MSWI), does not normalize the filters and has a
  looser relation with mean-square travel time error. If substantial
  energy is spread amongst multiple arrivals, on the other hand,
  neither AWI nor MSWI objectives approximate the travel time mean
  square erorr, and attempts to minimize them by local
  (Newton-related) optimization are as likely to stagnate at models
  predicting erroneous travel times as is least-square Full Waveform
  Inversion (FWI). The matching filter at the heart of this approach
  must be ``pre-whitened'', that is, computed by solving a regularized
  least squares problem for filtered data misfit. In order for the
  relation to traveltime tomography to hold, and for the filtered data
  misfit to stay relatively small, the regularization weight must be
  coupled to data wavelength. These objective functions can also be
  used as penalty terms in a homotopy of objectives connecting AWI or
  MSWI with FWI. \end{abstract}

\section{Introduction}
Full waveform inversion (FWI) refers to the inference of mechanical
parameter distribution within a material object  (the earth, a human
body, a manufactured artifact,...) from remote measurements of waves
propagating through it. These mechanical parameters appear as
coefficients in systems of partial differential equations governing
wave propagation, and numerical solutions of these wave equations can
be used to predict experimental data {\em ab initio}. FWI consists in
updating the parameters to minimize a numerical measure of  the difference between computationally
predicted wave data and actual observed
data. Such data is typically oscillatory. If wave velocities or their parametric equivalents are amongst the parameters
to be estimated, then updates cause shifts in the arrival times of
data oscillations, which in turn causes commonly used numerical error measures
to oscillate. This ``cycle skipping'' phenomenon may cause iterative local optimization algorithms (relatives of
Newton;'s method, the only practical algorithmic choices) to stagnate
far from good data fit, unless the initial
parameter estimates accurately predict the arrival times of waves \cite[]{GauTarVir:86,VirieuxOperto:09,HuetalIEEE:18}.

@@

This paper supplies such an explanation for a special case of AWI, the
recovery of smooth acoustic medium parameters from waves generated at
localized energy sources, propagated through a fluid, and recorded at
receiver locations remote from the sources (transmission data). I show that the AWI
objective function is an asymptotic (in time frequency) approximation
to the mean-square time-of-travel misfit, provided that arrival times
of transmitted waves are single. That is, a unique ray of geometric
asymptotics connects each sources-receiver pair, as is the case if the associated
Riemannian metric is simple in a domain containing the sources and
receivers \cite[]{PestovUhlmann:05,StefanovUhlmann:05}. The error in
this approximation is proportional to the source mean-square
wavelength. I also show that this relation fails if arrival times have
higher multiplicity. Travel time tomography (minimization of
mean-square travel time error) appears to be much less sensitive to
choice of initial model than is FWI. In fact, it is commonly used to
construct an initial model for FWI
\cite[]{Bordingetal:87,SirguePratt:04,VirieuxOperto:09}. At least for
acoustic transmission data with single arrivals, AWI implicitly
combines travel time tomography with FWI.

@@

  \begin{itemize}
  \item[1. ] $Tu_{\sigma}[\kappa;\dl](\bx_r,\cdot;\bx_s) \in
    L^2(\bR)$;
  \item[2. ] There exist $R_*,R^*,\lambda_*$ with $0<R_*\le R^*, 0 <
    \lambda_*\le 1$ depending on
    $M, X_{sr}$ so that
    \begin{itemize}
    \item[a. ] $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$ implies that $\sigma \le R^*\lambda$;
    \item[b. ] $\sigma \le R_*\lambda$ and $\lambda \le \lambda_*$
      implies that $\|\Sl[\kappa]\ul[\kappa;\dl]-\dl\| \le
      \frac{1}{2}\|\dl\|$.
    \end{itemize}
  \end{itemize}
  Choose $r > 0$ and assume that
  \begin{equation}
    \label{eqn:lamsig}
    \sigma = r \lambda.
  \end{equation}.
  Then there exist $0 < C_* \le C^*$, $0 < C$, and $W: M \times X_{sr}
  \rightarrow \bR^+$ with $C_* \le
  W \le C^*$ so that for
  $\log \kappa \in M$, $(\bx_s,\bx_r) \in X_{sr}$,
  \begin{itemize}
  \item[2. ] for $0 <  \lambda \le 1$, 
    \begin{equation}
      \label{eqn:mswitt}
      |\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|- \lambda^{-\frac{1}{2}} W[\kappa^*,\kappa]^{\frac{1}{2}}(\bx_s,\bx_r)|\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|| \le C
      \lambda^{\frac{1}{2}};
    \end{equation}
  \item[3. ]for sufficiently small $\lambda$, 
    \begin{equation}
      \label{eqn:awitt}
      \left|\frac{\|T\ul[\kappa;\dl](\bx_r,\cdot;\bx_s)\|}{\|\ul[\kappa,\dl](\bx_r,\cdot;\bx_s)\|}-
        |\tau[\kappa](\bx_s,\bx_r)-\tau[\kappa^*](\bx_s,\bx_r)|\right| \le C \lambda
    \end{equation}
  \end{itemize}
\end{theorem}
\noindent{\bf Remark:} The heuristic justification for AWI is the possibility of keeping the
filtered predicted data $\Fl[\kappa]*\ul[\kappa;d]$ close to the data
$d$ while updating the bulk modulus $\kappa$. The conclusions 2a and
2b show that for this to be the case, the regularization parameter $\sigma$
must be dominated by the wavelength proxy $\lambda$. This observation underlies the choice
$\sigma = r\lambda$ used in conclusions 3 and 4. A looser relation
would clearly suffice, however the argument lis slightly cleaner if
$\sigma$ is simply held proportional to $\lambda$ as $\lambda
\rightarrow 0$. Note that the units of $r$ and $\sigma$ are pressure$^2 \times$time$^2$.

Summation over the source-receiver positions in $X_{sr}$ yields
\begin{cor}
  \label{thm:cor1}
  In addition to the assumptions of Theorem \ref{thm:main}, define $J_{\rm
    AWI,\sigma}$ and $J_{\rm MSWI,\sigma}$ by equations
  \ref{eqn:awidefmain}, \ref{eqn:mswidefmain} respectively. Then
\begin{equation}
\label{eqn:cor1mswi}   
\lim_{\lambda \rightarrow 0} \lambda J_{\rm MSWI,r\lambda}[\kappa;\dl]
= \sum_{(\bx_s,\bx_r) \in X_{sr}} W[\kappa^*,\kappa](\bx_s,\bx_r) |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
\end{equation}
and
\begin{equation}
\label{eqn:cor1awi}   
\lim_{\lambda \rightarrow 0} J_{\rm AWI,r\lambda}[\kappa;\dl]
= \sum_{(\bx_s,\bx_r) \in X_{sr}}  |\tau[\kappa](\bx_s,\bx_r) - \tau[\kappa^*](\bx_s,\bx_r)|^2.
\end{equation}
\end{cor}

@@

@@

Of courese, the adaptive filter just computed is not square
integrable, and must be regularized for use in a square-integral
construction like that of AWI or MSWI. That amounts essentially to
replacing the $\delta$s in the above formulae by a square-integrable
approximate Dirac function, of roughly a wavelength in width. In the
preceding section, this approximate Dirac was $g$, so use the same
notation here. The regularized version of $u(t)$ (solution of the
least squares problem \ref{eqn:reg}) contains summands with multiples of $g(t-(\tau^*_1-\tau_0))$
and $g(t-(\tau_1-\tau^*_0))$. These are approximately scaled by
$\tau^*_1-\tau_0$ and $\tau_1-\tau^*_0$ respectively in the
definitions \ref{eqn:mswidefmain} and \ref{eqn:awidefmain} of the MSWI and AWI
objectives. For other choices of objective, they contribute similarly
to the total. They will eventually begin to cancel as $\kappa$ approaches
$\kappa^*$, however this cancellation happens only when the times are
within the width of the approximate 
Dirac waveform $g$ - that is, within a wavelength.

@@

%The reader will
%note that a normalization component of the AWI objective function
%definition is important in reaching this conclusion. The connection
%between objective normalization and the asymptotic equivalence to
%travel time inversion for transmission data has nothing to do with the
%reasons for normalizing the objective previoiusly advanced in the
%literature \cite[]{Warner:16,LiAlkhalifah:21}.

This paper has explained the behaviour of AWI (and MSWI) applied to
acoustic transmission data. For data exhibiting a single smooth
wavefront, these formulatios produce objective function that converge
to versions of travel time mean-square error, in the limit of hgh
data frequency. For

@@

Some of these constraints
could certainly be relaxed. For instance, exponential decay of local
energy implies that the finite time duration of actual measured data
can be treated as a perturbation of the infinite time data setting
posed here. The conclusions of this paper likely hold in some form
even for non-smooth sets of models, provided that the data for such models is
dominated by ballistic transmitted waves. The exception is the single-arrival hypothesis: it is
essential, as without it, the link between the travel time and
extended waveform inversion is lost.

This disconnect is a fearure of the adaptive filter construction, not of a particular objective function
(MSWI, AWI,...) formulated in terms of it. The introduction of the
adaptive filter $u$ in the filtered data prediction $F[\kappa]*u$ is equivalent
to permitting the source wavelet to depend on the receiver location,
which in the basic physical model of point source acoustics it does
not. Therefore this construction {\em extends} basic physical modeling
of acoustic wave transmission. \cite{HuangSymes:SEG15} call this is the {\em source-receiver
  extension}.  This ``crosstalk'' ambiguity analyzed in the previous
section afflicts all approaches based on the source-receiver
extension.

A number of authors have presented numerical illustrations of the
divergence between source-receiver extension methods and traveltime
tomography, and some have proposed remedies. \cite{Symes:94c}
illustrated the crosstalk phenomenon for a version of MSWI. Crosswell
source-receiver geometry often generates multiple arrivals,
particularly near low-velocity sedimentary layers that are natural
targets of seismic exploration and act as wave guides.
\cite{Plessix:00} successfully applied a version of MSWI to crosswell
waveform inversion by removing traces with close source and receiver
depths, thus leaving only data that tended to be dominated by single
arrivals. \cite{HuangSymes:Geo17} and \cite{SymesChenMinkoff:24}
illustrate the failure of the version of MSWI reported here to avoid
cycle-skipping in inversion of strongly refracting models, using
frequency-domain implementation of the basic definition
\ref{eqn:mswidefmain} and time-domain implementation of the reduced
penalty function \ref{eqn:mswipenred}, respectively.
\cite{Yongetal:GJI23} have shown that precisely the same issue arises
in AWI, and have proposed a remedy by localizing the definition of the
objective to short time intervals, within which multiple wavefront
arrrivals are unlikely to appear. \cite{Symes:23} constructs an
extension based on surface sources, as opposed to the point source -
point receiver extension which is the basis of MSWI and AWI. This
surface source extension separates events by infinitesimal slope,
information not available to the source-receiver extension.

Many open questions remain about MSWI, AWI, and other extension-based
approaches to transmitted wave data inversion. In particular, 
the analysis presented here has nothing to say about AWI applied to
reflected wave data. \cite{Warner:16} and others have
suggested that AWI is effective in that context. If so, the underlying
mechanism remains to be explained. The application of extended model
inversion to of elastodynamic transmission data remains to be investigated.
